---
title: "Task 1: Palmetto Binary Logistic Regression"
author: "Nick McManus"
date: "2023-02-12"
output: 
 html_document: 
    toc: yes
    toc_float: yes
    theme: cerulean
    code_folding: show
    smooth_scroll: yes
    collapsed: yes
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)  #always
library(GGally)     #plotting 
library(AICcmodavg) #compare AIC values
library(tidymodels) #k-fold cross validation
library(kableExtra) #tables
library(cowplot)    #figures
```

## Introduction
***

In this task, we create and compare two different binary logistic regression models for predicting palmetto species in south-central Florida. These models are created and tested using a long-term dataset of the two dominant palmetto species at the Archbold Biological Station: *Serenoa repens* and *Sabel etonia*. These data were collected at five-year intervals between 1981-2017.

![img/s_repens.jpeg]

To understand how well measured variables -- such as plant height, canopy length and width, and number of new green leaves -- can classify palmetto species, we'll first visualize the relationship between species and predictor variables. Then, metrics such as AICc, BIC, and cross validation are used to determine which of two models best predicts the palmetto species. 

**Source:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5

<br><br>

## Visual exporation
***

First we'll read in the data:
```{r}
palmetto <- read_csv("palmetto.csv") %>% 
  # keep variables of interest
  select(species, height:green_lvs) %>% 
  # change number to species name
  mutate(species = case_when(species == 1 ~ 'S. repens',
                             species == 2 ~ 'S. etonia'),
         species = factor(species)) %>% 
  # drop any NAs
  drop_na()

# Quickly visualize variable relationships
#ggpairs(data = palmetto, aes(color = species))
```

Then we'll visualize differences in height, canopy length, and the number of green leaves between palmetto species.
```{r,fig.height= 4, fig.width= 3.5}
## create overlapping plot
height_length_overlap <- ggplot(data = palmetto, aes(x = height, y = length)) +
  geom_point(aes(color = species)) +
  labs(x = 'Height (cm)',
       y = 'Canopy length (cm)') +
  scale_color_manual(values = c('seagreen4', 'seagreen2'))+
  theme_bw() +
  theme(
    #customize legend elements
    legend.position = c(0.15, 0.85),
    legend.title = element_text(face = 'bold', size = 9),
    legend.text = element_text(face = 'italic', size = 9),
    legend.background = element_rect(fill = 'honeydew'),
    legend.key = element_rect(fill = 'honeydew'),
    legend.key.size = unit(1, 'line'),
    legend.box.background = element_rect(color = 'black'),
    #customize axis elements
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_text(size = 9, face = 'bold'),
    axis.title.y = element_text(size = 9, face = 'bold'),
    panel.background = element_rect(fill = 'azure')) +
  #change legend dot size and title
  guides(colour = guide_legend(override.aes = list(size = 3), title = "Species:"))


## create facet wrapped plot
height_length_sep <- ggplot(data = palmetto, aes(x = height, y = length)) +
  geom_point(aes(color = species), alpha = 0.8, show.legend = FALSE) +
  labs(x = 'Height (cm)',
       y = 'Canopy length (cm)') +
  facet_wrap(~species) +
  scale_color_manual(values = c('seagreen4', 'seagreen2'))+
  theme_bw()+
  theme(
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_text(size = 9, face = 'bold'),
    axis.title.y = element_text(size = 9, face = 'bold'),
    panel.background = element_rect(fill = 'azure'))


## plot both graphs together
plot_grid(height_length_overlap, height_length_sep,
          align = 'V',
          ncol = 1,
          labels = c('A', 'B'),
          hjust = -1,
          vjust = 1.8)

```

**Figure 1.** Height (cm) and canopy length (cm) of Palmetto species measured at the Archbold Biological Station from 1981-2017. Observations for *S. etonia* and *S. repens* are displayed in dark and light green, respectively. Figure 1A directly compares observations for both species, while 1B separates observations to better visualize the spread of each species. 




```{r}
# histogram of green leaves by species
ggplot(data = palmetto, aes(x = green_lvs)) +
  #boxes for s.etonia
  geom_histogram(data = subset(palmetto, species == 'S. etonia'), 
                 aes(fill = species),
                 binwidth = 1, 
                 boundary = -0.5, #adjust axis for better labeling
                 color = 'black', 
                 alpha = 0.8) +
  #boxes for s.repens
  geom_histogram(data = subset(palmetto, species == 'S. repens'), 
                 aes(fill = species),
                 binwidth = 1, boundary = -0.5,
                 color = 'gray20',
                 alpha = 0.6) +
  #set colors and legend values
  scale_fill_manual(name = 'Species', 
                    values = c('seagreen4', 'seagreen2'),
                    labels = c('S. etonia', 'S. repens')) +
  #force origin to (0,0) and x-axis labels through 17
  scale_x_continuous(expand = c(0, 0), limits = c(0, 18), breaks=1:17) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 2100)) +
  #axis labels
  labs(x = 'Number of green leaves',
       y = 'Count') +
  #set theme
  theme_classic() +
  theme(
    axis.title.x = element_text(face = 'bold', vjust = -0.5),
    axis.title.y = element_text(face = 'bold'),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(vjust = 2),
    legend.position = c(0.8, 0.8),
    legend.title = element_text(face = 'bold', size = 10),
    legend.text = element_text(face = 'italic'),
    legend.key.size = unit(1, 'line'))

leaf_stats <- palmetto %>% 
  group_by(species) %>% 
  summarize(mean_lvs = mean(green_lvs),
            sd_lvs = sd(green_lvs))
```
**Figure 2.** The number of green leaves on a palmetto for two species measured at the Archbold Biological Station. Observations for *S. etonia* and *S. repens* are displayed in dark and light green, respectively. On average, *S. etonia* have `r round(leaf_stats$mean_lvs[1],1)` <font size="1">$\pm$</font> `r round(leaf_stats$sd_lvs[1],1)` green leaves while *S. repens* have `r round(leaf_stats$mean_lvs[2],1)` <font size="1">$\pm$</font> `r round(leaf_stats$sd_lvs[2],1)` green leaves (mean <font size="1">$\pm$</font> 1 standard deviation). The histograms' difference in spread and location reflect *S. repens* greater variability and average number of green leaves per plant.  




## Logistic regression

Create the models using tidy models
```{r}
# first formula considers all variables
f1 <- species ~ height + width + length + green_lvs
blr1 <- glm(formula = f1, data = palmetto, family = "binomial")

# second formula doesn't look at canopy length
f2 <- species ~ height + width + green_lvs
blr2 <- glm(f2, palmetto, family = "binomial")

# Get a tidy version w/ broom, then output nice tables
blr1_tidy <- broom::tidy(blr1)
blr2_tidy <- broom::tidy(blr2)

blr1_tidy %>% 
  kable(caption = "**Table 1.** Summary statistics for BLR model 1",
        col.names = c("Variable", "Estimate", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")

blr2_tidy %>% 
  kable(caption = "**Table 2.** Summary statistics for BLR model s",
        col.names = c("Variable", "Estimate", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
blr1_tidy
```


AICc and BIC values
```{r}
## make nice AICc table
aic_table <- aictab(list(blr1, blr2)) %>% 
  kable(caption = "**Table 3.** AICc values for Models 1 and 2.",
        col.names = c("Model", "Parameters", "AICc", 
                      "Delta AICc", "mdlLik","aiccwt", 
                      "Log Likelihood", "cum.wt")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left") %>% 
  remove_column(c(5,6,8)) #remove unnecessary columns
aic_table

## make nice BIC table
bic_table <- bictab(list(blr1, blr2)) %>% 
  kable(caption = "**Table 4.** BIC values for Models 1 and 2.",
        col.names = c("Model", "Parameters", "AICc", 
                      "Delta AICc", "mdlLik","aiccwt", 
                      "Log Likelihood", "cum.wt")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left") %>% 
  remove_column(c(5,6,8)) #remove unnecessary columns
bic_table
```


10-fold cross validation
```{r}
# set seed for reproducibility 
set.seed(123)

# set number of folds to 10 and repeat it 5 times
n_folds <- vfold_cv(palmetto, v = 10, repeats = 5)

# create BLR model
blr_model <- logistic_reg() %>% 
  set_engine('glm')



# set up workflow bundling logistic model with formula
## Run for formula 1 ----------------------
blr_tidy_wf1 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f1)

blr_tidy_cv_f1 <- blr_tidy_wf1 %>%
  fit_resamples(n_folds)

## output useful metrics
cv_metrics_f1 <- collect_metrics(blr_tidy_cv_f1)


## Run for forumula 2 ----------------------
blr_tidy_wf2 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f2)

blr_tidy_cv_f2 <- blr_tidy_wf2 %>%
  fit_resamples(n_folds)

## output useful metrics
cv_metrics_f2 <- collect_metrics(blr_tidy_cv_f2) 
 

# create tables for cv metrics ------------------
## model 1
cv_metrics_f1 %>% 
  kable(caption = '**Table X.** Cross validation metrics for Model 1',
        col.names = c('Metric', 'Estimator', 'Mean', 'n', 'Standard error', 'Configuration')) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
## model 2
cv_metrics_f2 %>% 
  kable(caption = '**Table X.** Cross validation metrics for Model 1',
        col.names = c('Metric', 'Estimator', 'Mean', 'n', 'Standard error', 'Configuration')) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")

```

Accuracy of Model 1 is `r round(cv_metrics_f1$mean[1],3)`.



Now train our selected Model 1 with the entire dataset and return coefficients. 
```{r}
# use the entire dataset to train model
blr1_tidyfit <- blr_model %>% 
  fit(f1, data = palmetto)

# output coefficients in nice table
coeffs_blr1_tidy <- tidy(blr1_tidyfit)
coeffs_blr1_tidy %>% 
  kable(caption = "**Table X.** Summary statistics for BLR model 1",
        col.names = c("Variable", "Estimate", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
```



```{r}
# convert log odds to probabilities with type.predict 
blr1_fitted <- blr1 %>%
  broom::augment(type.predict = 'response')

# function to calculate percent accuracy
pred_acc <- function(x, y) {
  accurate <- ifelse(x == y, 1, 0)
  return(mean(accurate, na.rm = TRUE))
}

# check the accuracy of the model by species
blr1_accuracy <- blr1_fitted %>% 
  #predict species based on 50% threshold
  mutate(species_pred = ifelse(.fitted > 0.5, 'S. repens', 'S. etonia')) %>% 
  #return 1 if prediction was correct, 0 if not
  mutate(correct_pred = ifelse(species_pred == species, 'yes', 'no')) %>% 
  #summary stats for both species, how many correct preds (and percentage)
  group_by(species) %>% 
  summarize(n_correct = sum(correct_pred == 'yes'),
            n_incorrect = sum(correct_pred == 'no'),
            pct_correct = pred_acc(species, species_pred))
  
# output nice table
blr1_accuracy %>% 
  kable(caption = "**Table X.** Accuracy...",
        col.names = c("Species", "Correct", "Incorrect", "Percent correct")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
  
```





Not sure if I still need this info...
```{r}

blr_f1_pred <- palmetto %>%
  mutate(predict(blr_tidyfit_f1, .),
         predict(blr_tidyfit_f1, ., type = 'prob'))

blr_f1_pred %>%
  roc_curve(truth = species, `.pred_S. etonia`) %>%
  autoplot()

blr_f1_pred %>%
  roc_auc(truth = species, `.pred_S. etonia`)
```



old purrr map code that didn't work:
```{r, eval = FALSE}
# function to calculate accuracy, given a "truth" vector and "prediction" vector
pred_acc <- function(x, y) {
  accurate <- ifelse(x == y, 1, 0)
  
  return(mean(accurate, na.rm = TRUE))
}

# function to calculate accuracy of BLR of one fold (training and testing)
calc_fold <- function(i, fold_df, f) {
  kfold_test <- fold_df %>%
    filter(fold == i)
  kfold_train <- fold_df %>%
    filter(fold != i)
  
  kfold_blr <- glm(f, data = kfold_train, family = 'binomial')
  kfold_pred <- kfold_test %>%
    mutate(blr = predict(kfold_blr, kfold_test, type = 'response')) %>%
    mutate(pred = ifelse(blr > 0.50, 'S. repens', 'S. etonia'))
  
  kfold_accuracy <- kfold_pred %>%
    summarize(blr_acc = pred_acc(species, pred)) # using my other function
  
  return(kfold_accuracy)
}


set.seed(24)

n_folds <- 10
fold_vec <- rep(1:n_folds, length.out = nrow(palmetto))

palmetto_kfold <- palmetto %>% 
  mutate(fold = sample(fold_vec, size = n(), replace = FALSE))


## run with purrr

results1 <- purrr::map(.x = 1:n_folds, # sequence of fold numbers
                       .f = calc_fold, # function
                       fold_df = palmetto_kfold, # additional argument to calc_fold()
                       f = f1) %>%              # additional argument to calc_fold()
  bind_rows() %>%
  mutate(mdl = 'f1')

results2 <- purrr::map(.x = 1:n_folds, .f = calc_fold, 
                               fold_df = palmetto_kfold,
                               f = f2) %>%
  bind_rows() %>%
  mutate(mdl = 'f2')

results_purrr_df <- bind_rows(results1, results2) %>%
  group_by(mdl) %>%
  summarize(mean_acc = mean(blr_acc))

results_purrr_df
```


























