---
title: "Task 1: Palmetto Binary Logistic Regression"
author: "Nick McManus"
date: "2023-02-12"
output: 
 html_document: 
    toc: yes
    toc_float: yes
    theme: cerulean
    code_folding: show
    smooth_scroll: yes
    collapsed: yes
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)  #always
library(GGally)     #plotting 
library(AICcmodavg) #compare AIC values
library(tidymodels) #k-fold cross validation
library(kableExtra) #tables
```

## Introduction
***

This analysis focuses on a dataset of south-central Florida palmetto species. This dataset includes (x, y, z) on the two dominant palmetto species at the Archbold Biological Station: *Serenoa repens* and *Sabel etonia*. Data was collected at five-year intervals between 1981-2017.

To understand how well measured variables -- such as plant height, canopy length and width, and number of new green leaves -- can classify palmetto species, we'll first visualize the relationship between species and predictor variables. Then, a binary logistic regression determines the probabilty of a given palmetto to be either species....




**Source:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5

<br><br>

## Visual exporation
***

First we'll read in the data:
```{r}
palmetto <- read_csv("palmetto.csv") %>% 
  # keep variables of interest
  select(species, height:green_lvs) %>% 
  # change number to species name
  mutate(species = case_when(species == 1 ~ 'S. repens',
                             species == 2 ~ 'S. etonia'),
         species = factor(species)) %>% 
  drop_na()

#ggpairs(data = palmetto, aes(color = species))
```

Then create visualizations.
```{r}
ggplot(data = palmetto, aes(x = height, y = length))+
  geom_point(aes(color = species), show.legend = FALSE) +
  labs(x = 'Height (cm)',
       y = 'Canopy Length (cm)') +
 # facet_wrap(~species) +
  scale_color_manual(values = c('seagreen4', 'seagreen2'))+
  theme()

## maybe make a top figure of them together, then bottom half of them split up and put it together in a cowplot? 
```
**Figure 1.** Height (cm) and canopy length (cm) of Palmetto species measured at the Archbold Biological Station from 1981-2017. Observations for *S. etonia* and *S. repens* are displayed in dark and light green, respectively. 




```{r}
# histogram of green leaves by species
ggplot(data = palmetto, aes(x = green_lvs, fill = species)) +
  geom_histogram() +
  labs(x = 'Number of Green Leaves',
       y = 'Count') +
  scale_fill_manual(values = c('seagreen4', 'seagreen2'))
```





## Logistic regression

Create the models using tidy models
```{r}
# first formula considers all variables
f1 <- species ~ height + width + length + green_lvs
blr1 <- glm(formula = f1, data = palmetto, family = "binomial")

# second formula doesn't look at canopy length
f2 <- species ~ height + width + green_lvs
blr2 <- glm(f2, palmetto, family = "binomial")

# Get a tidy version w/ broom, then output nice tables
blr1_tidy <- broom::tidy(blr1)
blr2_tidy <- broom::tidy(blr2)

blr1_tidy %>% 
  kable(caption = "**Table 1.** Summary statistics for BLR model 1",
        col.names = c("Variable", "Estimate", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")

blr2_tidy %>% 
  kable(caption = "**Table 2.** Summary statistics for BLR model s",
        col.names = c("Variable", "Estimate", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
blr1_tidy
```


AICc and BIC values
```{r}
## make nice AICc table
aic_table <- aictab(list(blr1, blr2)) %>% 
  kable(caption = "**Table 3.** AICc values for Models 1 and 2.",
        col.names = c("Model", "Parameters", "AICc", 
                      "Delta AICc", "mdlLik","aiccwt", 
                      "Log Likelihood", "cum.wt")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left") %>% 
  remove_column(c(5,6,8)) #remove unnecessary columns
aic_table

## make nice BIC table
bic_table <- bictab(list(blr1, blr2)) %>% 
  kable(caption = "**Table 4.** BIC values for Models 1 and 2.",
        col.names = c("Model", "Parameters", "AICc", 
                      "Delta AICc", "mdlLik","aiccwt", 
                      "Log Likelihood", "cum.wt")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left") %>% 
  remove_column(c(5,6,8)) #remove unnecessary columns
bic_table
```


10-fold cross validation
```{r}
# set seed for reproducibility 
set.seed(123)

# set number of folds to 10 and repeat it 5 times
n_folds <- vfold_cv(palmetto, v = 10, repeats = 5)

# create BLR model
blr_model <- logistic_reg() %>% 
  set_engine('glm')



# set up workflow bundling logistic model with formula
## Run for formula 1 ----------------------
blr_tidy_wf1 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f1)

blr_tidy_cv_f1 <- blr_tidy_wf1 %>%
  fit_resamples(n_folds)

## output useful metrics
cv_metrics_f1 <- collect_metrics(blr_tidy_cv_f1)


## Run for forumula 2 ----------------------
blr_tidy_wf2 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f2)

blr_tidy_cv_f2 <- blr_tidy_wf2 %>%
  fit_resamples(n_folds)

## output useful metrics
cv_metrics_f2 <- collect_metrics(blr_tidy_cv_f2) 
 

# create tables for cv metrics ------------------
## model 1
cv_metrics_f1 %>% 
  kable(caption = '**Table X.** Cross validation metrics for Model 1',
        col.names = c('Metric', 'Estimator', 'Mean', 'n', 'Standard error', 'Configuration')) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
## model 2
cv_metrics_f2 %>% 
  kable(caption = '**Table X.** Cross validation metrics for Model 1',
        col.names = c('Metric', 'Estimator', 'Mean', 'n', 'Standard error', 'Configuration')) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")

```

Accuracy of Model 1 is `r round(cv_metrics_f1$mean[1],3)`.



Now train our selected Model 1 with the entire dataset and return coefficients. 
```{r}
# use the entire dataset to train model
blr1_tidyfit <- blr_model %>% 
  fit(f1, data = palmetto)

# output coefficients in nice table
coeffs_blr1_tidy <- tidy(blr1_tidyfit)
coeffs_blr1_tidy %>% 
  kable(caption = "**Table X.** Summary statistics for BLR model 1",
        col.names = c("Variable", "Estimate", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
```



```{r}
# convert log odds to probabilities with type.predict 
blr1_fitted <- blr1 %>%
  broom::augment(type.predict = 'response')

# function to calculate percent accuracy
pred_acc <- function(x, y) {
  accurate <- ifelse(x == y, 1, 0)
  return(mean(accurate, na.rm = TRUE))
}

# check the accuracy of the model by species
blr1_accuracy <- blr1_fitted %>% 
  #predict species based on 50% threshold
  mutate(species_pred = ifelse(.fitted > 0.5, 'S. repens', 'S. etonia')) %>% 
  #return 1 if prediction was correct, 0 if not
  mutate(correct_pred = ifelse(species_pred == species, 'yes', 'no')) %>% 
  #summary stats for both species, how many correct preds (and percentage)
  group_by(species) %>% 
  summarize(n_correct = sum(correct_pred == 'yes'),
            n_incorrect = sum(correct_pred == 'no'),
            pct_correct = pred_acc(species, species_pred))
  
# output nice table
blr1_accuracy %>% 
  kable(caption = "**Table X.** Accuracy...",
        col.names = c("Species", "Correct", "Incorrect", "Percent correct")) %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = "hover",
                position = "left")
  
```





Not sure if I still need this info...
```{r}

blr_f1_pred <- palmetto %>%
  mutate(predict(blr_tidyfit_f1, .),
         predict(blr_tidyfit_f1, ., type = 'prob'))

blr_f1_pred %>%
  roc_curve(truth = species, `.pred_S. etonia`) %>%
  autoplot()

blr_f1_pred %>%
  roc_auc(truth = species, `.pred_S. etonia`)
```



old purrr map code that didn't work:
```{r, eval = FALSE}
# function to calculate accuracy, given a "truth" vector and "prediction" vector
pred_acc <- function(x, y) {
  accurate <- ifelse(x == y, 1, 0)
  
  return(mean(accurate, na.rm = TRUE))
}

# function to calculate accuracy of BLR of one fold (training and testing)
calc_fold <- function(i, fold_df, f) {
  kfold_test <- fold_df %>%
    filter(fold == i)
  kfold_train <- fold_df %>%
    filter(fold != i)
  
  kfold_blr <- glm(f, data = kfold_train, family = 'binomial')
  kfold_pred <- kfold_test %>%
    mutate(blr = predict(kfold_blr, kfold_test, type = 'response')) %>%
    mutate(pred = ifelse(blr > 0.50, 'S. repens', 'S. etonia'))
  
  kfold_accuracy <- kfold_pred %>%
    summarize(blr_acc = pred_acc(species, pred)) # using my other function
  
  return(kfold_accuracy)
}


set.seed(24)

n_folds <- 10
fold_vec <- rep(1:n_folds, length.out = nrow(palmetto))

palmetto_kfold <- palmetto %>% 
  mutate(fold = sample(fold_vec, size = n(), replace = FALSE))


## run with purrr

results1 <- purrr::map(.x = 1:n_folds, # sequence of fold numbers
                       .f = calc_fold, # function
                       fold_df = palmetto_kfold, # additional argument to calc_fold()
                       f = f1) %>%              # additional argument to calc_fold()
  bind_rows() %>%
  mutate(mdl = 'f1')

results2 <- purrr::map(.x = 1:n_folds, .f = calc_fold, 
                               fold_df = palmetto_kfold,
                               f = f2) %>%
  bind_rows() %>%
  mutate(mdl = 'f2')

results_purrr_df <- bind_rows(results1, results2) %>%
  group_by(mdl) %>%
  summarize(mean_acc = mean(blr_acc))

results_purrr_df
```


























